Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding Across Toronto’s Neighborhoods

### 2024-11-28 paper/paper.qmd
Update paper

## Overview
Urban biodiversity plays a crucial role in sustaining ecosystem health, yet cities face challenges in designing equitable and effective strategies to protect pollinators. Toronto’s PollinateTO initiative seeks to address this by funding community-led projects that create and enhance publicly visible pollinator gardens. Since its inception in 2019, the program has supported over 190 projects, resulting in approximately 500 gardens and 25,500 m² of pollinator habitat. While these efforts align with the city’s Pollinator Protection and Biodiversity Strategies, the allocation patterns of these grants remain an open question. Specifically, understanding whether certain types of gardens (e.g., rain gardens, boulevard gardens) or locations (e.g., Neighbourhood Improvement Areas) are more likely to receive funding can provide valuable insights into optimizing resources and maximizing program impact.

## Estimand
To address this question, this paper investigates the relationship between garden characteristics and PollinateTO funding decisions. Our analysis focuses on how variables such as garden type, location, and community engagement strategies influence both the frequency and magnitude of funding allocations. The estimand centers on whether specific garden features are associated with higher funding likelihoods and amounts, filling a critical gap in understanding how urban biodiversity programs distribute resources.

## Results
Our results reveal that certain garden types, such as_______, are more likely to secure funding, particularly in underserved areas identified as _______. We also find that projects emphasizing ______ receive higher allocations, highlighting the program’s alignment with broader engagement and ecological priorities. These findings offer actionable insights into how urban biodiversity initiatives can balance ecological goals with social equity considerations.

## Significance
The implications of this study extend beyond Toronto, providing a framework for other cities seeking to implement or refine pollinator protection strategies. By shedding light on funding allocation patterns, this research informs policymakers and urban planners on how to enhance the effectiveness and equity of biodiversity programs. These insights contribute to the growing body of literature on sustainable urban development and community-driven ecological stewardship.

## Telegraphing
The remainder of this paper is structured as follows. @sec-data describes the data used in the analysis, including its sources, key features, and considerations regarding measurement and visualization. Section 3 outlines the methodology, detailing the data visualization techniques employed and the justification for using a Bayesian logistic regression model to analyze the relationship between garden characteristics and funding outcomes. Section 4 presents the results, including model validation and key findings. Section 5 offers a discussion of the results, focusing on their implications through key points, followed by an examination of the study's limitations and suggested future steps for further exploration. Section 6 concludes the paper. The Appendix provides supplementary materials, including raw data previews, data feature definitions, data visualizations, and technical model details such as posterior predictive checks, MCMC convergence diagnostics, and additional figures.

# Data {#sec-data}
## Software Packages
We use the statistical programming language R [@citeR], rstanarm [@rstanarm], tidyverse [@tidyverse] Our data [@PTdata].... Following @tellingstories, we consider...

## Data Source
This study uses data from the PollinateTO Primary Project Garden Locations dataset, hosted on Toronto's Open Data Portal. The dataset provides information on community-led garden projects funded by PollinateTO, including geographic location, garden type, funding year, and other features relevant to understanding the distribution and characteristics of these initiatives. It is publicly accessible and serves as a vital resource for analyzing PollinateTO’s impact in Toronto neighborhoods.

## Data Features
The dataset includes several key variables:

YEAR_FUNDED: The year the project received funding from PollinateTO.
GARDEN_TYPE: Categories of gardens (e.g., food gardens, pollinator gardens, rain gardens).
NIA_OR_EN: Classification of garden location as part of a Neighbourhood Improvement Area (NIA), Emerging Neighborhood (EN), or other areas.
WARD_NAME: The city ward where the garden is located.
IS_INDIGENOUS_GARDEN: A binary variable indicating whether the garden is Indigenous-led.
PROJECT_NAME: Names assigned to specific garden projects.
LATITUDE and LONGITUDE: Geographic coordinates of each garden.
These features allow for diverse analyses, from exploring spatial trends to assessing equity and accessibility in funding distribution.

## Data Measurement*** (check key)
The dataset translates real-world community garden projects into structured entries by standardizing key attributes. For example:

YEAR_FUNDED records when projects were initiated, reflecting the temporal dimension of PollinateTO’s funding priorities.
GARDEN_TYPE is based on project applications, categorized to highlight garden purposes and ecological roles.
NIA_OR_EN and WARD_NAME capture the geographic and socioeconomic contexts of gardens, derived from city planning frameworks.
IS_INDIGENOUS_GARDEN identifies projects promoting Indigenous cultural or ecological knowledge, based on project self-identification during funding application.
The dataset provides structured snapshots of projects, enabling analysis of broader trends, equity impacts, and ecological contributions of urban greening initiatives.

@@

Self-Reported Data: Much of the data, including garden types and Indigenous-led status, relies on self-reporting by project applicants.
Selection Bias: Funding decisions prioritize certain areas (e.g., NIAs), potentially overrepresenting them compared to other regions.
Temporal Incompleteness: The dataset spans 2019 to present, limiting analyses to recent funding trends.
Granularity: Certain variables, such as project outcomes or specific plant species, are not included, restricting the scope of ecological impact analysis.
Geographic Precision: While coordinates are provided, some garden boundaries and precise sizes may not be captured, introducing potential spatial analysis challenges.

## Methodology
Several considerations affect the dataset’s completeness and reliability:

Self-Reported Data: Much of the data, including garden types and Indigenous-led status, relies on self-reporting by project applicants.
Selection Bias: Funding decisions prioritize certain areas (e.g., NIAs), potentially overrepresenting them compared to other regions.
Temporal Incompleteness: The dataset spans 2019 to present, limiting analyses to recent funding trends.
Granularity: Certain variables, such as project outcomes or specific plant species, are not included, restricting the scope of ecological impact analysis.
Geographic Precision: While coordinates are provided, some garden boundaries and precise sizes may not be captured, introducing potential spatial analysis challenges.

@@

# Use the log-transformed column in the boxplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, notch = FALSE) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
# Model
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project.

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

>>>>

## Overview
Urban biodiversity is increasingly recognized as vital to maintaining healthy ecosystems, especially in the face of rapid urbanization. Among the most important urban biodiversity goals is the protection of pollinators, whose decline poses significant risks to food security, agricultural productivity, and ecological stability. In response to this, many cities have launched initiatives to create pollinator-friendly habitats within urban spaces. Toronto’s PollinateTO program is one such initiative, aiming to enhance biodiversity by funding community-led pollinator gardens. Since its launch in 2019, the program has supported over 190 projects, resulting in nearly 500 gardens and approximately 25,500 square meters of pollinator habitat.

Despite the program's success in expanding green spaces, the allocation of funding across these projects remains an understudied area. Questions about which types of gardens or neighborhoods are more likely to receive funding, and the factors influencing these decisions, are crucial for optimizing the program’s impact. The objective of this paper is to examine these allocation patterns, specifically whether garden characteristics such as type (e.g., rain gardens, boulevard gardens) or location (e.g., Neighbourhood Improvement Areas) influence the likelihood of receiving funding and the size of the funding allocation.

In particular, this paper addresses the gap in understanding how urban biodiversity funding decisions are made, with a focus on the specific factors that contribute to higher funding likelihoods and amounts. The estimand in this analysis is whether particular garden features, community engagement strategies, or geographic locations are associated with receiving more substantial funding allocations from the PollinateTO program. By identifying these key factors, this research provides important insights into the effectiveness and equity of funding distributions for urban biodiversity projects.

Our findings indicate that garden type, community engagement, and the location of the project all play significant roles in funding decisions. Specifically, we find that gardens in underserved neighborhoods and those emphasizing community involvement are more likely to receive higher funding allocations. These results offer actionable insights for optimizing resource distribution in urban biodiversity programs and aligning ecological goals with social equity considerations.

The remainder of the paper is structured as follows: @sec-data describes the data used in the analysis, including its sources, key features, and considerations regarding measurement and visualization. Section 3 outlines the methodology, detailing the data visualization techniques employed and the justification for using a Bayesian logistic regression model to analyze the relationship between garden characteristics and funding outcomes. Section 4 presents the results, including model validation and key findings. Section 5 offers a discussion of the results, focusing on their implications through key points, followed by an examination of the study's limitations and suggested future steps for further exploration. Section 6 concludes the paper. The Appendix provides supplementary materials, including raw data previews, data feature definitions, data visualizations, and technical model details such as posterior predictive checks, MCMC convergence diagnostics, and additional figures.

# Data {#sec-data}
## Software Packages
This analysis utilizes the R programming language [@citeR], with the 'rstanarm' [@rstanarm] and 'tidyverse' [@tidyverse] packages for statistical modeling and data manipulation. Our data comes from the PollinateTO Primary Project Garden Locations dataset [@PTdata], which is hosted on Toronto's Open Data Portal. Following the approach outlined in @tellingstories, we leverage these tools to explore patterns and trends in the allocation of PollinateTO funding.

## Data Source
The primary source of data for this study is the PollinateTO Primary Project Garden Locations dataset, which contains detailed information on community-led garden projects funded by PollinateTO. The dataset includes key features such as project locations, garden types, funding years, and other relevant information that enables an in-depth analysis of funding patterns and their spatial and socio-economic implications. This dataset is publicly accessible through Toronto's Open Data Portal and is a vital resource for studying the impact and effectiveness of PollinateTO in promoting urban biodiversity.

## Data Features
The dataset contains several key variables that are critical for understanding the distribution and characteristics of funded garden projects. Below, we describe each variable and its relevance to the analysis:

- YEAR_FUNDED: The year each project received funding from PollinateTO. This variable is crucial for understanding temporal trends in the program’s funding allocation. It also allows us to analyze shifts in funding priorities over time.
- GARDEN_TYPE: This variable categorizes the gardens by their ecological functions (e.g., rain gardens, food gardens, pollinator gardens). These types reflect the intended ecological role of each garden and are important for understanding the program's impact on biodiversity and the environment.
- NIA_OR_EN: This classification indicates whether the project is located in a Neighbourhood Improvement Area (NIA), Emerging Neighbourhood (EN), or another area. These labels are derived from Toronto's planning data and allow for an exploration of whether funding is disproportionately allocated to areas with different socio-economic characteristics, enabling us to analyze equity in the distribution of funds.
- WARD_NAME: The specific ward in Toronto where the garden project is located. This geographic indicator helps us assess whether certain neighborhoods receive more funding based on local government priorities or demographic characteristics.
- IS_INDIGENOUS_GARDEN: A binary variable that identifies whether the project is Indigenous-led, based on self-identification in the funding application. This is a key variable for exploring whether the program is aligned with Toronto’s commitments to Indigenous communities and ecological knowledge.
PROJECT_NAME: The name assigned to each garden project. While this variable is useful for identification, it is not directly involved in the analysis but helps in linking data points to specific projects.
- LATITUDE and LONGITUDE: These geographic coordinates provide spatial context for each garden project. While these data points help in mapping and visualizing the geographic distribution of funded gardens, their primary role is in understanding spatial patterns and localizing project locations.

These features collectively allow for a broad analysis of trends, equity, and the ecological goals of the PollinateTO initiative.

## Constructed Variables
In this study, variable is_large_garden was constructed. This binary variable was constructed to classify gardens as "large" or "small" based on their estimated garden size. Specifically, gardens with an estimated garden size greater than the mean size of all gardens in the dataset were coded as 1 (large garden), while those with a size less than or equal to the mean were coded as 0 (small garden). The mean garden size was calculated using the estimated_garden_size variable, excluding any missing values. This classification was chosen to capture differences in garden size relative to the overall distribution and to investigate whether larger gardens are more likely to receive funding.

## Data Measurement
The dataset translates real-world community garden projects funded by PollinateTO into structured entries, enabling the analysis of patterns and relationships. It includes several key variables, with the following interpretations based on assumptions about the dataset’s construction:

YEAR_FUNDED likely reflects administrative records of the year projects received funding.
GARDEN_TYPE seems derived from categorizations in project applications, capturing intended ecological roles.
NIA_OR_EN appears to use Toronto city planning designations to classify locations.
IS_INDIGENOUS_GARDEN assumes self-identification during the funding application process.
WARD_NAME is based on project location within Toronto’s municipal wards.

These assumptions are informed by typical practices in grant-based datasets. Confirmation would require consultation with PollinateTO’s documentation or administrators.

@@

Self-Reported Data: Variables such as GARDEN_TYPE and IS_INDIGENOUS_GARDEN are based on self-reporting by project applicants, which may introduce some subjectivity and inconsistency.

Selection Bias: The dataset reflects funding decisions that may prioritize certain geographic areas, such as NIAs and ENs, which could lead to an overrepresentation of these areas relative to others. This potential bias should be considered when analyzing equity and resource allocation.

Temporal Incompleteness: The dataset covers the period from 2019 to the present, which limits our ability to analyze trends over a longer historical period.

Granularity: Some ecological measures, such as specific plant species or detailed environmental outcomes, are not captured in the dataset. This restricts the scope of ecological impact analysis, though the dataset’s focus on garden characteristics and funding allocation provides valuable insights.

Geographic Precision: While LATITUDE and LONGITUDE provide the coordinates for each garden, there may be some inaccuracies in the precise location or boundaries of gardens, particularly in cases where they span multiple properties or are irregularly shaped.

## Methodology
The PollinateTO dataset collects detailed information on community-led garden projects funded by Toronto’s PollinateTO initiative. Data is sourced from the city’s Open Data portal and includes variables like garden type, location, funding year, and size, enabling analysis of trends in urban biodiversity projects. The dataset uses standardized attributes and is updated annually, making it a valuable tool for evaluating the initiative’s impact across various Toronto neighborhoods, though it may be subject to self-reporting biases and geographical limitations in precision.

@@

# Use the log-transformed column in the boxplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, notch = FALSE) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
# Model
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project. We are interested in modeling whether a garden is large (binary outcome) as a function of various factors such as the garden type, year funded, neighborhood type, indigenous garden status, and ward name. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

```{Latex}
\begin{align}
y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha + \beta_1 \times \mbox{garden_type}_i + \beta_2 \times \mbox{year_funded}_i + \beta_3 \times \mbox{nia_or_en}_i + \beta_4 \times \mbox{is_indigenous_garden}_i + \beta_5 \times \mbox{ward_name}_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\beta_4 &\sim \mbox{Normal}(0, 2.5) \\
\beta_5 &\sim \mbox{Normal}(0, 2.5)
\end{align}
```

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@
